Embedding Dimension of Polyhedral Losses

### 2020-01-31 colt-20/embed-dim.tex
writing in section 3

In this section, we completely characterize when a discrete loss can be embedded into the real line.
For example, this will immediately imply (see Section \ref{sec:examples}) that the abstain losses on $n \geq 3$ outcomes require $d \geq 2$.
We will also give a constructive algorithm for testing $1$-embeddability and constructing a $1$-dimensional polyhedral surrogate.
Finally, we show that $1$-embeddability is equivalent to the existence of \emph{any} convex calibrated $1$-dimensional surrogate.
After presenting and discussing this sequence of results, we observe that they can be collected as a set of five or six conditions on $\ell$ that are all pairwise equivalent, and in particular, are equivalent to $1$-embeddability.

A condition from the literature~\citep{lambert2018elicitation} turns out to be key to $1$-dimensional embeddings.
\begin{definition}[Orderable]\label{def:orderable-hyperplane}
	A finite property $\gamma:\simplex \to 2^\R$ is \emph{orderable} if there is an enumeration of $\R = \{r_1, \ldots, r_{|\R|}\}$ such that for all $i \leq |\R| - 1$, we have $\gamma_{r_i} \cap \gamma_{r_{i+1}}$ is a hyperplane intersected with $\simplex$.
\end{definition}
This definition follows from~\cite[Theorem 3]{lambert2018elicitation}.

Before investigating the usefulness of this condition, we give an equivalent condition which may be more intuitive.
Given a finite property $\gamma$, elicited by a discrete loss $\ell$, a common technique particularly in information elicitation is to visualize the level sets as a power diagram (generalization of Voronoi diagram) in the simplex\bo{cite}.
See Figures \ref{fig:intersection-graph-ex} and \ref{fig:abstain-alpha-half}.
We can formalize the adjacency relationships of these level sets, giving a visual test for orderability.

\begin{definition}[Intersection graph]\label{def:intersection-graph}
  Given a discrete loss $\ell$ and associated finite property $\gamma: \simplex \to 2^{\R}$ elicited by $\ell$, the \emph{intersection graph} has vertices labeled by $\R$ with an edge $(r,r')$ if $\gamma_r \cap \gamma_{r'} \cap \relint(\simplex) \neq \emptyset$, where $\relint(\simplex)$ equals the set of distributions with full support.
\end{definition}
\begin{proposition} \label{prop:orderable-iff-path}
  A finite property $\gamma$ is \emph{orderable} if and only if its intersection graph is a path, i.e. a connected graph where every vertex has either one or two neighbors.
\end{proposition}
\begin{proof}
  $(\implies)$The intersection graph is constructed by adding an edge for each halfspace, which connects only two nodes.
  If three level sets intersected on $\relint(\simplex)$, then the level boundary for any two would not be a halfspace (this follows because the level sets form a power diagram, e.g. \bo{cite}).
  This yields a path for the intersection graph.

  \bigskip
  \bo{This one needs a bit of work.}
  $(\impliedby)$ If the intersection graph forms a path, then each level set has at most two neighbors.
  The level sets are convex polytopes, as they form a power diagram\dots.

@@

@@

Our first main result is the following; we defer discussion of its proof to the end of the section.
\begin{theorem} \label{thm:orderable-iff-1d}
  A discrete loss $\ell$ is $1$-embeddable if and only if the property it elicits is orderable.
\end{theorem}
In other words, to know whether a one-dimensional polyhedral surrogate for $\ell$ exists, it suffices to check whether the intersection graph is a path.
We give an example of a direct application in Section \ref{sec:examples}.

\subsection{Constructing a surrogate}
Theorem \ref{thm:orderable-iff-1d} is only somewhat constructive.
Next, we give a more constructive test for $1$-embeddability.
In fact, in the event that $\ell$ is $1$-embeddable, we will actually build a $1$-dimensional polyhedral calibrated surrogate.

@@

  Moreover, when these conditions hold, the loss $L:\reals\to\reals^\Y$ embeds $\ell$ with $\varphi:\R\to\reals$, \jessiet{Have been using $\phi$ instead of $\varphi$ for embeddings in this paper so far. I have a slight preference for $\varphi$ since it matches the last paper, but I don't care too much... just need to pick one.}\raft{Why switch to $\phi$ if you prefer $\varphi$?}
  where
  \begin{align*}
    \varphi(r_i) &= \sum_{j=1}^{i-1} 1/\Lambda_j~, \quad \text{(where $\varphi(r_1)=0$)}
    \\
    L(u) &= \begin{cases}
      \ell(r_1) - u K \ones & u \leq \varphi(r_1) = 0 \\
      \ell(r_i) + \Lambda_i \cdot (u - \varphi(r_i)) \cdot  (\ell(r_{i+1})-\ell(r_i)) & u \in [\varphi(r_i),\varphi(r_{i+1})] \\
      \ell(r_k) + \Lambda_{k-1} \cdot (u - \varphi(r_k))\cdot K \ones & u \geq \varphi(r_k)
    \end{cases}~,
  \end{align*}
  \jessiet{In the cases: what if $\varphi(r_1) < 0$?}\raft{I set $\varphi(r_1)=0$; updated to something hopefully clearer}
  where $\lambda(i) = \min(\lambda^+(i),\max(\lambda^-(i),1))$, $\Lambda_i := \prod_{j=2}^i \lambda(j)$, $\Lambda_1=1$, and $K = \max_{i\in\{2,\ldots,k\},y\in\Y} |v(i)_y|$.
\end{theorem}
Note that a calibrated surrogate should also specify a link function $\psi: \reals^d \to \R$.
We have been omitting it because of the guarantee of Theorem \ref{thm:embed-iff-poly}, from \cite{finocchiaro2019embedding}, which guarantees its existence.
But in the case of $d=1$, we observe that one can simply use the midpoints between the embedding points as cutoffs for the link function, i.e. $\psi(u) = \argmin_{r \in \R} |u - \phi(r)|$, breaking ties arbitrarily.

\subsection{Proof techniques and general convex surrogates}
We collect the above results, and add one more, in the following summary theorem.
\begin{theorem} \label{thm:1d-tfae}
  Let $\ell: \R \to \nonnegreals^{\Y}$ be a discrete loss eliciting a finite property $\gamma$.
  The following are equivalent: (1) $\gamma$ is orderable; (2) the intersection graph of $\gamma$ is a path; (3) the two conditions of Theorem \ref{thm:construct-1d-loss} are satisfied; (4) $\ell$ is $1$-embeddable; (5) $\ell$ has some polyhedral calibrated surrogate loss $L: \reals \to \nonnegreals^{\Y}$; (6) $\ell$ has some convex calibrated surrogate loss $L: \reals \to \nonnegreals^{\Y}$.
\end{theorem}

@@

  Moreover, when these conditions hold, the loss $L:\reals\to\reals^\Y$ embeds $\ell$ with $\varphi:\R\to\reals$, \jessiet{Have been using $\phi$ instead of $\varphi$ for embeddings in this paper so far. I have a slight preference for $\varphi$ since it matches the last paper, but I don't care too much... just need to pick one.}\raft{Why switch to $\phi$ if you prefer $\varphi$?}
  where
  \begin{align*}
    \varphi(r_i) &= \sum_{j=2}^i 1/\Lambda_j~, \quad \text{(where $\varphi(r_1)=0$)}
    \\
    L(u) &= \begin{cases}
      \ell(r_1) - u K \ones & u \leq \varphi(r_1) = 0 \\
      \ell(r_i) + \Lambda_i \cdot (u - \varphi(r_i)) \cdot  (\ell(r_{i+1})-\ell(r_i)) & u \in [\varphi(r_i),\varphi(r_{i+1})] \\
      \ell(r_k) + \Lambda_{k-1} \cdot (u - \varphi(r_k))\cdot K \ones & u \geq \varphi(r_k)
    \end{cases}~,
  \end{align*}
  \jessiet{In the cases: what if $\varphi(r_1) < 0$?}\raft{I set $\varphi(r_1)=0$; updated to something hopefully clearer}
  where $\lambda(i) = \min(\lambda^+(i),\max(\lambda^-(i),1))$, $\Lambda_i := \prod_{j=2}^i \lambda(j)$, $\Lambda_1=1$, and $K = \max_{i\in\{2,\ldots,k\},y\in\Y} |v(i)_y|$.
\end{theorem}
\begin{proof}
  \raf{Actually, the easiest proof given what we have is:
    \\
  orderable
  $\implies$
  these $\ell$ conditions
  $\implies$
  1-embeddable
  $\implies$
  monotone
  $\implies$
  orderable
  \\
  Note: need $\gamma$ and $\Gamma$ in a few places}

>>>>

In this section, we completely characterize when a discrete loss $\ell$ can be embedded into the real line, i.e., when $\ell$ is 1-embeddable.
Our first characterization is expressed in terms of the property $\gamma$ that $\ell$ elicits, stating that $\ell$ is 1-embeddable if and only if $\gamma$ is \emph{orderable}, meaning the adjacency graph of its level sets is a path.
For example, this characterization will immediately imply that embedding the abstain losses on $n \geq 3$ outcomes requires $d \geq 2$ dimensions (\S~\ref{sec:examples}).
While determining these adjacencies can be straightforward when $\ell$ has known symmetries, we also give a more constructive algorithm for testing $1$-embeddability and constructing a $1$-dimensional polyhedral surrogate.
Finally, we show that the existence of any $1$-dimensional convex calibrated implies $1$-embeddability, showing that embeddings are without loss of generality in dimension 1.
After presenting and discussing this sequence of results, we observe that they can be collected as a set of six conditions on $\ell$ that are all pairwise equivalent, and in particular, are equivalent to $1$-embeddability.

\subsection{General characterization via property elicitation}

We begin with conditions on the property elicited by a discrete loss.
The following condition of~\citet[Theorem 3]{lambert2018elicitation}, that a finite property is \emph{orderable}, states that any two level sets intersect in a hyperplane, or not at all.
\begin{definition}[Orderable]\label{def:orderable-hyperplane}
	A finite property $\gamma:\simplex \to 2^\R$ is \emph{orderable} if there is an enumeration of $\R = \{r_1, \ldots, r_{|\R|}\}$ such that for all $i \leq |\R| - 1$, we have $\gamma_{r_i} \cap \gamma_{r_{i+1}}$ is a hyperplane intersected with $\simplex$.
\end{definition}

In fact, we show that orderability characterizes $1$-embeddability.
The proof involves several intermediate results which we state later in this section; see \S~\ref{sec:1d-proof-discuss} for more details. 
\begin{theorem} \label{thm:orderable-iff-1d}
  A discrete loss $\ell$ is $1$-embeddable if and only if the property it elicits is orderable.
\end{theorem}

We next give an equivalent condition to orderability which may be more intuitive: the adjacency graph of the level sets of $\gamma$, formed by connecting reports if their level sets intersect, must be a path.
This graph can be easily established for discrete losses with known symmetries or other facts, such as abstain, the mode, or ranking losses.

\begin{definition}[Intersection graph]\label{def:intersection-graph}
  Given a discrete loss $\ell$ and associated finite property $\gamma: \simplex \toto 2^{\R}$ elicited by $\ell$, the \emph{intersection graph} has vertices $\R$ with an edge $(r,r')$ if $\gamma_r \cap \gamma_{r'} \cap \relint(\simplex) \neq \emptyset$, where $\relint(\simplex)$ is the relative interior of $\simplex$.
  %(distributions with full support).
\end{definition}
\begin{proposition} \label{prop:orderable-iff-path}
  A finite property $\gamma$ is \emph{orderable} if and only if its intersection graph is a path, i.e. a connected graph where every vertex has either one or two neighbors.
\end{proposition}
\begin{proof}
  $(\implies)$
  The intersection graph is constructed by adding an edge for each halfspace, which connects only two nodes.
  If three level sets intersected on $\relint(\simplex)$, then the level boundary for any two would not be a halfspace (this follows because the level sets form a power diagram, e.g. \bo{cite}).
  This yields a path for the intersection graph.

  $(\impliedby)$
  \bo{This one needs a bit of work.}
  If the intersection graph forms a path, then each level set has at most two neighbors.
  The level sets are convex polytopes, as they form a power diagram\dots.

@@

Combining Proposition~\ref{prop:orderable-iff-path} and Theorem~\ref{thm:orderable-iff-1d}, we see that in order for $\ell$ to be embedded onto the real line, it is necessary and sufficient for the intersection graph of the property $\gamma$ to be a path.
We give an example of a direct application in Section \ref{sec:examples}.
By visualizing the level sets of $\gamma$ as a power diagram (generalization of Voronoi diagram) in the simplex~\cite{lambert2009eliciting}, we can also use Proposition~\ref{prop:orderable-iff-path} to perform a visual test for orderability, and thus 1-embeddability (Figures~\ref{fig:intersection-graph-ex} and~\ref{fig:abstain-alpha-half}).

@@

\subsection{Constructing a surrogate}
While Theorem~\ref{thm:orderable-iff-1d} and Proposition~\ref{prop:orderable-iff-path} are quite useful for discrete losses with known symmetries, they do not immediately provide an algorithm to test $1$-embeddability of an arbitrary discrete loss $\ell$, nor to construct the convex loss $L$ which embeds it.
We now turn to a more algorithmic test, which will actually build a real-valued polyhedral calibrated surrogate in the event that $\ell$ is $1$-embeddable.

@@

  Moreover, when these conditions hold, the loss $L:\reals\to\reals^\Y$ embeds $\ell$ with $\varphi:\R\to\reals$,
  where
  \begin{align*}
    \varphi(r_i) &= \sum_{j=1}^{i-1} 1/\Lambda_j~, \quad \text{(where $\varphi(r_1)=0$)}
    \\
    L(u) &= \begin{cases}
      \ell(r_1) - u K \ones & u \leq \varphi(r_1) = 0 \\
      \ell(r_i) + \Lambda_i \cdot (u - \varphi(r_i)) \cdot  (\ell(r_{i+1})-\ell(r_i)) & u \in [\varphi(r_i),\varphi(r_{i+1})] \\
      \ell(r_k) + \Lambda_{k-1} \cdot (u - \varphi(r_k))\cdot K \ones & u \geq \varphi(r_k)
    \end{cases}~,
  \end{align*}
  \jessiet{In the cases: what if $\varphi(r_1) < 0$?}\raft{I set $\varphi(r_1)=0$; updated to something hopefully clearer}
  where $\lambda(i) = \min(\lambda^+(i),\max(\lambda^-(i),1))$, $\Lambda_i := \prod_{j=2}^i \lambda(j)$, $\Lambda_1=1$, and $K = \max_{i\in\{2,\ldots,k\},y\in\Y} |v(i)_y|$.
\end{theorem}
As intuition for the proof, note that the conditions of the theorem ensure the existence of a positive multiplier $\lambda(i)$ making $v(i) \leq \lambda(i) v(i+1)$ hold coordinate-wise; our choice of $\lambda(i)$ is but one option.
The construction of $L$ sets the left and right derivatives at an embedding point $\varphi(r_i)$ to be positive multiples of $v(i)$ and $v(i+1)$, respectively, using this inequality to maintain monotonicity, and hence convexity of $L$.
The vectors $v(i),v(i+1)$ are chosen precisely to give the correct optimality conditions, so that for a given distribution, $r_i$ is optimal for $\ell$ if and only if $\varphi(r_i)$ is optimal for $L$.
The reverse direction, showing that these conditions are necessary for $1$-embeddability, is much more involved (\S~\ref{sec:1d-proof-discuss}).

Note that a calibrated surrogate should also specify a link function $\psi: \reals^d \to \R$.
As our surrogate is an embedding, Theorem \ref{thm:embed-iff-poly} guarantees the existence of a link achieving calibration.
We can also easily construct one in the case of $d=1$, by taking the midpoints between the embedding points as cutoffs: $\psi(u) = \argmin_{r \in \R} |u - \varphi(r)|$, breaking ties arbitrarily.

\subsection{Proof techniques and general convex surrogates}
\label{sec:1d-proof-discuss}

\raf{*much* more to say here...}

We collect the above results, and add one more, in the following summary theorem.
\begin{theorem} \label{thm:1d-tfae}
  Let $\ell: \R \to \nonnegreals^{\Y}$ be a discrete loss eliciting a finite property $\gamma$.
  The following are equivalent: (1) $\gamma$ is orderable; (2) the intersection graph of $\gamma$ is a path; (3) the two conditions of Theorem \ref{thm:construct-1d-loss} are satisfied; (4) $\ell$ is $1$-embeddable; (5) $\ell$ has some polyhedral calibrated surrogate loss $L: \reals \to \nonnegreals^{\Y}$; (6) $\ell$ has some convex calibrated surrogate loss $L: \reals \to \nonnegreals^{\Y}$.
\end{theorem}

@@

  Moreover, when these conditions hold, the loss $L:\reals\to\reals^\Y$ embeds $\ell$ with $\varphi:\R\to\reals$,
  where
  \begin{align*}
    \varphi(r_i) &= \sum_{j=2}^i 1/\Lambda_j~, \quad \text{(where $\varphi(r_1)=0$)}
    \\
    L(u) &= \begin{cases}
      \ell(r_1) - u K \ones & u \leq \varphi(r_1) = 0 \\
      \ell(r_i) + \Lambda_i \cdot (u - \varphi(r_i)) \cdot  (\ell(r_{i+1})-\ell(r_i)) & u \in [\varphi(r_i),\varphi(r_{i+1})] \\
      \ell(r_k) + \Lambda_{k-1} \cdot (u - \varphi(r_k))\cdot K \ones & u \geq \varphi(r_k)
    \end{cases}~,
  \end{align*}
  \jessiet{In the cases: what if $\varphi(r_1) < 0$?}\raft{I set $\varphi(r_1)=0$; updated to something hopefully clearer}
  where $\lambda(i) = \min(\lambda^+(i),\max(\lambda^-(i),1))$, $\Lambda_i := \prod_{j=2}^i \lambda(j)$, $\Lambda_1=1$, and $K = \max_{i\in\{2,\ldots,k\},y\in\Y} |v(i)_y|$.
\end{theorem}
\begin{proof}
  \raf{Actually, the easiest proof given what we have is:
    \\
  orderable
  $\implies$
  these $\ell$ conditions
  $\implies$
  1-embeddable
  $\implies$
  monotone
  $\implies$
  orderable
  \\
  Note: need $\gamma$ and $\Gamma$ in a few places}
